---
title: "Starwars"
author: "Bobr"
date: "21 04 2021"
output:
  html_document: default
---
## Setting up my environment
```{r loading packages}
library(tidyverse)
library(ggpubr)
library(ggpmisc)
```
## Cleaning the set
In this process we have done:

* removed heroes with a mass over 1000 kg.
* dropped heroes without gender.
* created dynamic variable for heroes quantity.
```{r cleaning the set}
starwars_c <- starwars %>%
  drop_na(sex) %>% 
  filter(mass < 1000) 
n_heroes <- nrow(starwars_c) #to make caption dinamic
head(starwars_c, n=6)
```
## Calculating statistical data
Running this code results in creating the following table:
```{r calculating statistical data}
starwars_stat <- starwars_c %>% 
  group_by(sex) %>% 
  summarise(mean(height), sd(height), mean(mass), sd(mass), cor=cor(height,mass)) 
head(starwars_stat)
```
## Plotting
Here we plot(scatter) heroes' height against body mass and look at the breakdown by gender.
```{r plotting}
ggplot(data = starwars_c,
       aes(x = height,
           y = mass)) +
  geom_point(aes(color = sex), size = 3) +
  stat_smooth(aes(color = sex, linetype=sex), method = "lm", se = F, formula = y ~ x) +
    geom_density2d_filled(alpha = 0.5) +
  stat_cor(aes(color = sex), digits=2, p.accuracy=0.01, method="pearson", size = 5) +
    labs(title = "SW: Body Mass vs. Height", subtitle = "Male - Female - Not Specified",
       caption = paste0("Based on ", n_heroes, " heroes"), x="Height",y="Weight") +
  annotate("text", x=215, y=72, label = "well stacked", color = "orange",
           fontface = "bold", size = 4.5, angle=-10)
```

Then we plot(bar) height against body mass, look at the breakdown by gender , and create a ribbon of plots by eye color.
```{r}
ggplot(data = starwars_c) +
  geom_bar(aes(x=sex, fill = eye_color)) + facet_wrap(~eye_color)
```

